---
title: "Untitled"
format: html
editor: visual
echo: false
---

## 

# Formal data analysis

```{r}
#| echo: false
#| eval: true
#| warning: false

library(ggplot2)
library(tidyverse)
library(gt)
library(tidymodels)
```

### Questions of interest

There are two questions of interest in this study:\
(a) Did Edinburgh exhibit a change in measles susceptibility following the retraction of the Wakefield's article?\
(b) Did the change, if any, in measles susceptibility occur in 2004 alongside the article's retraction?

```{r}
#| echo: false
#| eval: true

measles <- read.csv("DAProject14.csv")
```

### Methodology

First, the measles susceptibility rate for each of Edinburgh's 101 Intermediate Zones (IZ) was calculated using the formula:

$$
Measles~Susceptibility~Rate=\frac{Susceptibility~Children}{Total~Children}
$$

This computation yielded a new column, as presented in the following table.

```{r}
#| echo: false
#| eval: true
#| cache: true

# Calculate the measles susceptibility rate
measles$rate <- measles$Y / measles$N

rate_table <- measles %>%
  head() %>%
  gt() %>%
  tab_header(title = "Measles Susceptibility Data") %>%
  cols_label(
    Year = "Year",
    Y = "Susceptible Children",
    N = "Total Children",
    rate = "Susceptibility Rate") %>%
  tab_style(
    style = cell_text(align = "center"),
    locations = cells_body(columns = everything()))

rate_table
```

Next, the dataset was grouped by year to derive the yearly average measles susceptibility rate, as shown in the table below.

```{r}
#| echo: false
#| eval: true

# Calculate the yearly average measles susceptibility rate
yearly_avg_rate <-  measles %>%
  group_by(Year) %>%
  summarise(Average_Rate = mean(rate))

avg_rate_table <- yearly_avg_rate %>%
  head() %>%
  gt() %>%
  tab_header(title = "Yearly Average Measles Susceptibility Rate Data") %>%
  cols_label(
    Year = "Year", 
    Average_Rate = "Average Susceptibility Rate")

avg_rate_table
```

### Forming a model

Here we will fit a simple linear regression model to the `measles` data.

```{r}
#| eval: true
#| echo: false

model <- linear_reg() |>
  fit(Year~Average_Rate, data=yearly_avg_rate)
model$fit
```

Our best-fitting line to the data is:

$$
\widehat{average ~ rate}= \hat{\alpha} + \hat{\beta}x_i = 2023.3 - 132.5\cdot{year}
$$ where:

-   $\hat{\alpha}$=2023.3 is the intercept coefficient and means that for a given IZ with a `Year=0`, their `Average_Rate` would be 2023.3. Note that `Year=0` is not actually possible in our data set.

-   $\hat{\beta}$=-132.5 is the slope coefficient associated with the explanatory variable `Average_Rate`. This summarises the relationship between our response variable, `Year`, and our explanatory variable `Average_Rate`. Since the slope coefficient is negative, this means that `Year` and `Average_Rate` have a negative linear relationship, i.e. as `Average_Rate` increases, `Year` decreases. Furthermore, for every 1 unit increase in `Average_Rate`, there is an associated decrease of, on average 132.5 units of `Year`.

Here is the model summary.

```{r}
#|eval: true
#|echo: false
tidy(model)

```

@fig-avgprop-year-lm is a plot of our data with a line of best fit superimposed.

```{r}
#| eval: true
#| label: fig-avgprop-year-lm
#| fig-cap: Jittered scatterplot of measles susceptibility by year

ggplot(yearly_avg_rate, aes(x=Year, y=Average_Rate)) +
  geom_jitter() +
  labs(x= "Year", y="Average proportion of measle susceptibility", title="Measles susceptibility by year") +
  geom_smooth(method="lm", se=FALSE)

```

### Visualization and Analysis

A line graph was plotted based on the yearly average measles susceptibility rates, as presented in the figure below.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-align: center
#| label: fig-line
#| fig-cap: Line graph of measles susceptibility rate in Edinburgh (1998-2014)

# Plot a line graph
ggplot(yearly_avg_rate, aes(x = Year, y = Average_Rate)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue",size = 3) +
  geom_vline(xintercept = 2004, color = "red", linetype = "dashed") +
  geom_text(aes(x = 2004, y = min(yearly_avg_rate$Average_Rate), label = "2004"),
            color = "red", size = 4) +
  labs(title = "Measles Susceptibility Rate in Edinburgh (1998-2014)",
       x = "Year", y = "Average Susceptibility Rate")
```

The graph clearly indicates that the yearly average measles susceptibility rate initially increased, peaked in 2004, and then consistently declined.

Furthermore, a t-test was conducted to statistically evaluate the difference in measles susceptibility rates before and after 2004.

$$
H_0:\mu_{pre2004}=\mu_{post2004}~~~~\text{and}~~~~H_1:\mu_{pre2004}\neq\mu_{post2004}
$$

The table shows that p-value is less than 0.05, so we reject the null hypothesis.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-align: center

# Split the data (with 2004 as the boundary)
pre_2004 <- measles %>% 
  dplyr::filter(Year < 2004) %>% 
  pull(rate)
post_2004 <- measles %>% 
  dplyr::filter(Year >= 2004) %>% 
  pull(rate)

# t test
t_test <- t.test(pre_2004, post_2004)

t_test_df <- data.frame(
  "Statistic" = c("t-value", "Degrees of Freedom", "p-value", 
                  "95% CI Lower", "95% CI Upper", 
                  "Mean of pre_2004", "Mean of post_2004"),
  "Value" = c(
    round(t_test$statistic, 4),
    round(t_test$parameter, 4),
    round(t_test$p.value, 4),
    round(t_test$conf.int[1], 4),
    round(t_test$conf.int[2], 4),
    round(t_test$estimate[1], 4),
    round(t_test$estimate[2], 4)
  ))

t_test_table <- t_test_df %>%
  gt() %>%
  tab_header(title = "T-Test Results") %>%
  cols_label(
    Statistic = "Statistic",
    Value = "Value")

t_test_table
```

### Conclusion

These findings answer the second question, providing evidence to support that the observed change in measles susceptibility rates is associated with the retraction of the Wakefield's article.
