---
title: "Group Project"
format: html
editor: visual
execute: 
  echo: false
  warning: false
  message: false
---

# 1. Introduction

Between 1998 and 2014, data were collected by The Scottish Childhood Immunisation Record System on measles, mumps and rubella (MMR) vaccination uptake. In 1998, Wakefield et al. published an article which linked the MMR vaccine with an increased risk of autism, which decreased the vaccination rates in 2003 across the United Kingdom. The article was partially withdrawn in 2004, then later discredited in 2010. This report aims to answer two main questions:

-   Did Glasgow exhibit a change in measles susceptibility following the retraction of the Wakefield article?

-   Did the change, if any, in measles susceptibility occur in 2004 alongside the articles’ retraction?

# 2. Exploratory Analysis

We begin our exploratory analysis by creating a new column in the dataset which calculates the proportion of pre-school children that are susceptible to measles in each of Edinburgh's 101 Intermediate Zones (IZ). This is what we are interested in modelling. For each year, we can calculate the mean proportion of children susceptible to measles. A summary is presented below:

```{r}
#| include: false
library(ggplot2)
library(tidyverse)
library(gt)
library(tidymodels)
library(skimr)
library(moderndive)
library(dplyr)
```

```{r}
#| include: false
data <- read.csv("DAProject14.csv")

```

```{r}
#| label: tbl-summary
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Mean proportion of pre-school children susceptible to measles by year."

data$prop <- data$Y/data$N
yearly_avg <- data %>%
  group_by(Year) %>%
  summarise(Mean_Susceptibility = mean(prop))
yearly_avg_table <-  pivot_wider(yearly_avg, names_from = "Year", values_from = "Mean_Susceptibility")
yearly_avg_table |> gt() |> 
  fmt_number(decimals=3)

```

From @tbl-summary, we can see that between 1998 and 2004, the mean proportion of pre-school children susceptible to measles increased. With a larger increase between 2002 and 2004. After 2004, there was a sharp decrease in the mean proportions. The mean proportion of pre-school children susceptible to measles in 2014 is approximately half of the mean proportion in 2006.

```{r}
#| echo: false
#| fig-cap: Distribution of mean proportion of measles susceptibility by year (1998 - 2014). 
#| label: fig-boxplot
#| fig-align: center
#| fig-height: 3.5
#| fig-width: 4.5
ggplot(data, aes(x = as.factor(Year), y = prop)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(x = "Year",
       y = "Proportion Susceptible") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

From @fig-boxplot, we can see that the variability in the proportion of children susceptible is greater in the earlier years (1998-2004), since the boxplots are wider. The median proportion increased between 1998 and 2004, then began to decrease from 2006 onwards. Most of the boxplots appear to be symmetric, however there are a few outliers for each year.

```{r}
#|echo: false
#|fig-cap: "Mean measles susceptibilty over time"
#|label: fig-scatterplot
#| fig-align: center
#| fig-width: 4.5
#| fig-height: 3.5

ggplot(yearly_avg, aes(x = Year, y = Mean_Susceptibility)) +
  geom_line(colour = "turquoise3", size = 1) +
  geom_point(colour = "palevioletred", size = 2) +
  geom_vline(xintercept = 2004, linetype = "dashed", colour = "black")
```

From @fig-scatterplot we can see that there is a peak in mean susceptibility in 2004 before a sharp decrease afterwards, this aligns with all other plots and summaries we have seen before.

```{r}
#|fig-cap: "Count of IZ with the same value of Y separated by Year"
#|label: fig-barY
#| fig-align: center
#| fig-width: 4.5
#| fig-height: 3.5

ggplot(data = data, aes(x = Y, fill = Year)) +
  geom_bar(show.legend = F) + 
  facet_wrap(Year~.) +
  scale_fill_binned()

```

The histograms for the years 1998-2004 look relatively similar. From 2006 onwards, the histograms appear to be slightly left-skewed. For the earlier years (1998-2004), there is more variability in the number of children susceptible to measles in each of Edinburgh's 101 Intermediate Zones (IZ).

# 3. Formal data analysis

```{r}
#| echo: false
#| eval: true
#| warning: false

library(ggplot2)
library(tidyverse)
library(gt)
library(tidymodels)
```

```{r}
#| echo: false
#| eval: true

measles <- read.csv("DAProject14.csv")
```

### 3.1 Methodology

First, the measles susceptibility rate for each of Edinburgh's 101 Intermediate Zones (IZ) was calculated using the formula:

$$
Measles~Susceptibility~Rate=\frac{Susceptibility~Children}{Total~Children}
$$

This computation yielded a new column, as presented in the following table.

```{r}
#| echo: false
#| eval: true
#| cache: true

# Calculate the measles susceptibility rate
measles$rate <- measles$Y / measles$N

rate_table <- measles %>%
  head() %>%
  gt() %>%
  tab_header(title = "Measles Susceptibility Data") %>%
  cols_label(
    Year = "Year",
    Y = "Susceptible Children",
    N = "Total Children",
    rate = "Susceptibility Rate") %>%
  tab_style(
    style = cell_text(align = "center"),
    locations = cells_body(columns = everything()))

rate_table
```

Next, the dataset was grouped by year to derive the yearly average measles susceptibility rate, as shown in the table below.

```{r}
#| echo: false
#| eval: true

# Calculate the yearly average measles susceptibility rate
yearly_avg_rate <-  measles %>%
  group_by(Year) %>%
  summarise(Average_Rate = mean(rate))

avg_rate_table <- yearly_avg_rate %>%
  head() %>%
  gt() %>%
  tab_header(title = "Yearly Average Measles Susceptibility Rate Data") %>%
  cols_label(
    Year = "Year", 
    Average_Rate = "Average Susceptibility Rate")

avg_rate_table
```

### 3.2 Forming a model

Here we will fit a simple linear regression model to the `measles` data.

```{r}
#| eval: true
#| echo: false

model <- linear_reg() |>
  fit(Year~Average_Rate, data=yearly_avg_rate)
model$fit
```

Our best-fitting line to the data is:

$$
\widehat{average ~ rate}= \hat{\alpha} + \hat{\beta}x_i = 2023.3 - 132.5\cdot{year}
$$ where:

-   $\hat{\alpha}$=2023.3 is the intercept coefficient and means that for a given IZ with a `Year=0`, their `Average_Rate` would be 2023.3. Note that `Year=0` is not actually possible in our data set.

-   $\hat{\beta}$=-132.5 is the slope coefficient associated with the explanatory variable `Average_Rate`. This summarises the relationship between our response variable, `Year`, and our explanatory variable `Average_Rate`. Since the slope coefficient is negative, this means that `Year` and `Average_Rate` have a negative linear relationship, i.e. as `Average_Rate` increases, `Year` decreases. Furthermore, for every 1 unit increase in `Average_Rate`, there is an associated decrease of, on average 132.5 units of `Year`.

Here is the model summary.

```{r}
#|eval: true
#|echo: false
tidy(model)

```

@fig-avgprop-year-lm is a plot of our data with a line of best fit superimposed.

```{r}
#| eval: true
#| label: fig-avgprop-year-lm
#| fig-cap: Jittered scatterplot of measles susceptibility by year

ggplot(yearly_avg_rate, aes(x=Year, y=Average_Rate)) +
  geom_jitter() +
  labs(x= "Year", y="Average proportion of measle susceptibility", title="Measles susceptibility by year") +
  geom_smooth(method="lm", se=FALSE)

```

### 3.3 Visualization and Analysis

A line graph was plotted based on the yearly average measles susceptibility rates, as presented in the figure below.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-align: center
#| label: fig-line
#| fig-cap: Line graph of measles susceptibility rate in Edinburgh (1998-2014)

# Plot a line graph
ggplot(yearly_avg_rate, aes(x = Year, y = Average_Rate)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue",size = 3) +
  geom_vline(xintercept = 2004, color = "red", linetype = "dashed") +
  geom_text(aes(x = 2004, y = min(yearly_avg_rate$Average_Rate), label = "2004"),
            color = "red", size = 4) +
  labs(title = "Measles Susceptibility Rate in Edinburgh (1998-2014)",
       x = "Year", y = "Average Susceptibility Rate")
```

The graph clearly indicates that the yearly average measles susceptibility rate initially increased, peaked in 2004, and then consistently declined.

Furthermore, a t-test was conducted to statistically evaluate the difference in measles susceptibility rates before and after 2004.

$$
H_0:\mu_{pre2004}=\mu_{post2004}~~~~\text{and}~~~~H_1:\mu_{pre2004}\neq\mu_{post2004}
$$

The table shows that p-value is less than 0.05, so we reject the null hypothesis.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-align: center

# Split the data (with 2004 as the boundary)
pre_2004 <- measles %>% 
  dplyr::filter(Year < 2004) %>% 
  pull(rate)
post_2004 <- measles %>% 
  dplyr::filter(Year >= 2004) %>% 
  pull(rate)

# t test
t_test <- t.test(pre_2004, post_2004)

t_test_df <- data.frame(
  "Statistic" = c("t-value", "Degrees of Freedom", "p-value", 
                  "95% CI Lower", "95% CI Upper", 
                  "Mean of pre_2004", "Mean of post_2004"),
  "Value" = c(
    round(t_test$statistic, 4),
    round(t_test$parameter, 4),
    round(t_test$p.value, 4),
    round(t_test$conf.int[1], 4),
    round(t_test$conf.int[2], 4),
    round(t_test$estimate[1], 4),
    round(t_test$estimate[2], 4)
  ))

t_test_table <- t_test_df %>%
  gt() %>%
  tab_header(title = "T-Test Results") %>%
  cols_label(
    Statistic = "Statistic",
    Value = "Value")

t_test_table
```

# 4. Conclusion

These findings answer the second question, providing evidence to support that the observed change in measles susceptibility rates is associated with the retraction of the Wakefield's article.
